Modelling the emergence of whisker barrels

### 2020-05-22 paper/briefpaper/brief.tex
The final tag in the repo will just be eLife
WIP on paper

\usepackage{amsmath,esint}
\newcommand*\dif{\mathop{}\!\mathrm{d}}

@@

% Blue for comments
\newcommand{\cmnt}[1]{\textcolor{blue}{#1}}
\newcommand{\dvrg}{\nabla\vcdot\nabla}
\newcommand{\e}{\emph}
\newcommand{\bol}{\textbf}
\newcommand{\mb}[1]{\mathbf{#1}}
\makeatletter
\newcommand*\vcdot{\mathpalette\vcdot@{.35}}
\newcommand*\vcdot@[2]{\mathbin{\vcenter{\hbox{\scalebox{#2}{$\m@th#1\bullet$}}}}}
\newcommand{\code}[1]{\textsf{#1}}

@@

\cite{karbowski_model_2004} developed a reaction-diffusion style model of how
extrinsic signalling gradients can constrain the emergence of distinct fields
from intrinsic cortical dynamics. Their model defines how the density of
connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a
1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by
$i$. The model was derived from the assumption that the rates at which $a_i$
and $c_i$ grow are reciprocally coupled. Extending the original 1D model to
simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and
$c_i(\mb{x},t)$, and model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise the connection density increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. Note that computing the divergence in
simulation requires cells on the cortical sheet to communicate with immedately
adjacent cells only (see \emph{Materials \& Methods}) The second term on the
right quantifies the coupling between axon branching and synaptogenesis. Here
  $\chi_i=0$ is a placeholder.

@@

\begin{figure}
\begin{fullwidth}
\includegraphics[width=\linewidth]{./MainFig.png}
\caption{\textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
  by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
  measure departure (Honda-$\delta$; see \citealp{senft_mouse_1991}) from a
  perfect Voronoi tesselation. Right shows the initial distribution of axon
  branching density ($a$) for one thalamocortical projection, and two
  molecular guidance fields ($\rho$). \textbf{B} The strengths of interaction
  $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41
  projections by the lengths of green and blue arrows respectively, assuming
  that similar fields aligned to the posterior-anterior and medial-lateral
  axes in the ventroposterior medial nucleus of the thalamus are sampled at
  the locations of putative barreloid centers (reconstructed from
  \citealp{haidarliu_size_2001}). \textbf{C} Simulation results for parameters
  $N=41$, $\alpha=3$, $\beta=20$, $k=3$, $D=0.2$, $\gamma\in\pm 2$,
  $\epsilon=150$ and $\delta{t}=0.0001$. Colours indicate the thalamic
  projection for which the connection density is maximal, black lines
  delineate boundaries, and overlaid contours show $c>0.5$ (see Movie
  S1). \textbf{D} Red dots show the Honda-$\delta$ metric obtained from
  simulation approaching that obtained from barrels in \textbf{A} (dotted
  line); black squares measure the correspondence between the real and
  simulated barrel shapes, $\eta$ (the product of the sum of squared
  differences between real and simulated centers and the sum of differences in
  area; units mm$^4$). \textbf{E} Guidance fields and emergent barrel pattern
  in a Fgf8 misexpression experiment
  (c.f. \citealp{assimacopoulos_fibroblast_2012}), simulated by reflecting
  $\rho_1$ at the join ($\epsilon=80$). All scale bars 1mm.}
\label{fig:main}
\end{fullwidth}
\end{figure}

@@

The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching we
redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other
projections are dense. Divisive normalization keeps the axon branch density
bounded on each iteration to the initial (random) density of unconnected
  branches at $t=0$
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \oiint_{S} \big[ a_i(\mathbf{x}, 0) - c_i(\mathbf{x}, t) \big] \dif S \; \frac {a'_i(\mathbf{x}, t)} {\oiint_{S}
  a'_i(\mathbf{x}, t) \dif S}\,,
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from
Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$ (${\delta}t$ is the time
represented by one simulation step). Note that this
  operation is local to individual afferent projections.

@@

All code required to reproduce these results is available at
\url{https://github.com/ABRG-Models/BarrelEmerge/tree/eLife_submission1}. The
computations described in (i), (ii) and (iii) may be found in the class method
\code{RD\_James::compute\_divJ()} which calculates \code{term1}, \code{term2}
and \code{term3}, respectively.

>>>>

\usepackage{color}
\usepackage{amsmath,esint}

% Some document-defined commands
\newcommand*\dif{\mathop{}\!\mathrm{d}}
\newcommand{\cmnt}[1]{\textcolor{blue}{#1}}
\newcommand{\dvrg}{\nabla\vcdot\nabla}
\newcommand{\e}{\emph}
\newcommand{\bol}{\textbf}
\newcommand{\mb}[1]{\mathbf{#1}}
\makeatletter
\newcommand*\vcdot{\mathpalette\vcdot@{.35}}
\newcommand*\vcdot@[2]{\mathbin{\vcenter{\hbox{\scalebox{#2}{$\m@th#1\bullet$}}}}}
\newcommand{\code}[1]{\textsf{#1}}

@@

@@

\cite{karbowski_model_2004} developed a reaction-diffusion style model of how
extrinsic signalling gradients can constrain the emergence of distinct fields
from intrinsic cortical dynamics. Their model defines how the density of
connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a
1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by
$i$. The model was derived from the assumption that the rates at which $a_i$
and $c_i$ grow are reciprocally coupled. Extending the original 1D model to
simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and
$c_i(\mb{x},t)$, and model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise the connection density increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x}) \cmnt{+ \chi_i}\right) - \frac{\partial c_i}{\partial t}.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. Note that computing the divergence in
simulation requires cells on the cortical sheet to communicate with immedately
adjacent cells only (see \emph{Materials \& Methods}) The second term on the
right quantifies the coupling between axon branching and synaptogenesis. Here
  $\chi_i=0$ is a placeholder.

@@

\begin{figure}
\begin{fullwidth}
%\includegraphics[width=\linewidth]{./MainFig.png}
\includegraphics[width=\linewidth]{./MainFig_superlowres.png}
\caption{\textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
  by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
  measure departure (Honda-$\delta$; see \citealp{senft_mouse_1991}) from a
  perfect Voronoi tesselation. Right shows the initial distribution of axon
  branching density ($a$) for one thalamocortical projection, and two
  molecular guidance fields ($\rho$). \textbf{B} The strengths of interaction
  $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41
  projections by the lengths of green and blue arrows respectively, assuming
  that similar fields aligned to the posterior-anterior and medial-lateral
  axes in the ventroposterior medial nucleus of the thalamus are sampled at
  the locations of putative barreloid centers (reconstructed from
  \citealp{haidarliu_size_2001}). \textbf{C} Simulation results for parameters
  $N=41$, $\alpha=3$, $\beta=20$, $k=3$, $D=0.2$, $\gamma\in\pm 2$,
  $\epsilon=150$ and $\delta{t}=0.0001$. Colours indicate the thalamic
  projection for which the connection density is maximal, black lines
  delineate boundaries, and overlaid contours show $c>0.5$ (see Movie
  S1). \textbf{D} Red dots show the Honda-$\delta$ metric obtained from
  simulation approaching that obtained from barrels in \textbf{A} (dotted
  line); black squares measure the correspondence between the real and
  simulated barrel shapes, $\eta$ (the product of the sum of squared
  differences between real and simulated centers and the sum of differences in
  area; units mm$^4$). \textbf{E} Guidance fields and emergent barrel pattern
  in a Fgf8 misexpression experiment
  (c.f. \citealp{assimacopoulos_fibroblast_2012}), simulated by reflecting
  $\rho_1$ at the join ($\epsilon=80$). All scale bars 1mm.}
\label{fig:main}
\end{fullwidth}
\end{figure}

@@

The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching we
% EITHER
%%redefined
%%%
%%\begin{equation} \label{eq:comp}
%%\color{blue}
%%\chi_i(\mb{x}, t) = \frac{\bar{a}_i F}{N-1} \nabla \sum_{j\ne i}{a_j}
%%\color{black}
%%\end{equation}
% OR
\cmnt{first define a quantity to represent the branching density of all
  \emph{other} TC types; $\hat{a}_i = \sum_{j\ne i}{a_j}$; then set the term
  $\chi_i$ to be}
%
\begin{equation} \label{eq:comp}
\color{blue}
\chi_i(\mb{x}, t) = \frac{\bar{a}_i F}{N-1} \nabla \hat{a}_i
\color{black}
\end{equation}

%
\cmnt{where $F$ is a parameter and $\bar{a}_i$ is a bounded version of $a_i$
  obtained by passing the branching density through the sigmoid function}
%
\begin{equation} \label{eq:abar}
\color{blue}
  \bar{a}_i = \frac{2}{1+\mathrm{e}^{5 - 0.5 a_i}}.
\color{black}
\end{equation}
%
\cmnt{In this expression for $\bar{a}_i$ (only), we model the idea
 that there must be an upper bound on the axonal branching density.}

\cmnt{which contributes to the \emph{flux of axonal branching} by making branching
  density for each projection `flow away' from regions} where the branches of
other projections are dense. Note that this operation is local to individual
afferent projections.

@@

All code required to reproduce these results is available at
\url{https://github.com/ABRG-Models/BarrelEmerge/tree/eLife}. The
computations described in (i), (ii) and (iii) may be found in the class method
\code{RD\_James::compute\_divJ()} which calculates \code{term1}, \code{term2}
and \code{term3}, respectively.
